LaMachine: A meta-distribution for NLP software

### 2019-03-29 docs/lamachine_clarin2018.tex
[paper] rephrased switchboard comparison and other parts to make clearer that LaMAchine is a distribution

Software is a key deliverable and a vital component for research in projects such as those under the CLARIN umbrella.
Software can be seen as an invaluable instrument that enable researchers to perform their studies. It is CLARIN's core mission to make
digital language resources, including software, available to the wider research community.

We see that NLP software often takes on complex forms such as processing pipelines invoking various individual
components, which in turn rely on various dependencies. Add dedicated web-interfaces on top of that and you obtain a
suite of interconnected software that is often non-trivial to install, configure, and deploy. This is where LaMachine
comes in.

@@

LaMachine incorporates software providing different types of interfaces which, as seen above, typically address different audiences. Whilst we
attempt to accommodate both technical\footnote{Data scientists, DevOps, system administrators, developers.} and
less-technical audiences\footnote{The wider researcher community, particularly the Humanities; also educational
settings.}, there is a natural bias towards the former as lower-level interfaces are often a prerequisite to build
higher-level interfaces on. Depending on the \emph{flavour} (more on this later) of LaMachine chosen, it makes a good virtual research
environment for a data scientist, whether on a personal computer or on a computing cluster, a good development
environment for a developer or a good deployment method for production servers in for example CLARIN centres.

@@

LaMachine exists since May 2015 and has been installed extensively ever since by numerous users. Quantifying its actual
use is not trivial despite having some tracking solutions in place, but our statistics show that LaMachine was installed at least over a thousand times, on
machines in over 25 different countries since its inception\footnote{https://applejack.science.ru.nl/lamastats/lamachinestats.html}. In early 2018 version 2 was released which was a significant redesign, powered
by Ansible\footnote{\url{https://www.ansible.com}}. We ourselves use LaMachine as a means by to make our full software
stack available to end-users, whether those be individual researchers, research groups, or hosting providers like CLARIN
centres. LaMachine has in fact been adopted in all of those areas. We also see a regular influx of questions on our issue tracker, indicating community interest.

@@

In LaMachine, software is grouped into various ``packages'', each package\footnote{For those familiar with
Ansible, a package in LaMachine is an Ansible role} groups one or multiple programs that have some kind of relation. The
installation manifest lists all packages that will be installed, at the user's discretion. After the initial
installation, the user can always add more packages using \texttt{lamachine-add} or editing the installation manifest directly.

@@

In any of its flavours, LaMachine offers low-level shell access, i.e. a command-line interface accessed through a
terminal. In flavours that are separated from the host system by a network, this is accomplished over \texttt{ssh}.
After accessing the LaMachine environment through a terminal, as shown in Figure~\ref{fig:venv}, the user has the liberty to do whatever she wants and can invoke any
of the tools that offer a text interface, including text editors such as vim, emacs or nano, version control systems
such as git and interpreters such as Python or R (or compilers for that matter). This allow users to use
one of the many specialised programming libraries included in LaMachine to build their own tools and . All this makes
LaMachine ideally suited as a development environment.

@@

LaMachine provides various CLAM-based webservices. CLAM is a long-time CLARIN project that allows developers to turn
their command-line NLP tool into a RESTful webservice with relatively little effort \cite{CLAMPAPER}, and which in
addition also provides a generic web-interface for human end-users. The portal website in LaMachine links to these. This
again addresses an audience that does not require a high degree of technical expertise, but is suitable for a wider
audience of researchers, including those in the Humanities.

@@

Within the scope of CLARIN, one might be inclined to draw comparisons with the CLARIN Language Resource Switchboard
\cite{zinn2016clarin}, which also offers access to NLP tools through a web interface where users can upload their texts
for NLP processing, providing easy access and automated tool discovery.  This, however, is notably different from what
LaMachine is. The CLARIN switchboard is not a software distribution, it links to externally hosted applications; and
LaMachine does currently not provide any switchboard functionality, but merely offers a portal interface. Moreover,
LaMachine is used by data scientists and reseachers who do have programming experience and want to work with complex NLP
software, on their own machines and unencumbered by high-level interfaces if they so desire.

There are also CLARIN initiatives that provide not just a switchboard, but a more elaborate high-level virtual research
environment (VRE) with graphical interfaces to accommodate the researcher and allow execution and visualisation of
various NLP pipeline. The Dutch CLARIAH WP3 project\footnote{\url{https://github.com/meertensinstituut/clariah-wp3-vre}}
is such an initiative and LaMachine is currently developed in close collaboration with this project, in a role of technology supplier.

>>>>

Software is a key deliverable and a vital component for research in projects such as those under the CLARIN umbrella.
Software can be seen as an invaluable instrument that enables researchers to perform their studies. It is CLARIN's core mission to make
digital language resources, including software, available to the wider research community.

We see that NLP software often takes on complex forms such as processing pipelines invoking various individual
components, which in turn rely on various dependencies. Add dedicated web-interfaces on top of that and you obtain a
suite of interconnected software that is often non-trivial to install, configure, and deploy. This is where LaMachine
comes in; as a \emph{distribution of software}.

@@

LaMachine distributes software providing different types of interfaces which, as seen above, typically address different audiences. Whilst we
attempt to accommodate both technical\footnote{Data scientists, DevOps, system administrators, developers.} and
less-technical audiences\footnote{The wider researcher community, particularly the Humanities; also educational
settings.}. There is a natural bias towards the former as lower-level interfaces are often a prerequisite to build
higher-level interfaces on. Depending on the \emph{flavour} (more on this later) of LaMachine chosen, it makes a good virtual research
environment for a data scientist, whether on a personal computer or on a computing cluster, a good development
environment for a developer or a good deployment method to deliver to production servers such as CLARIN centres.

@@

LaMachine exists since May 2015 and has been installed extensively ever since by numerous users. Quantifying its actual
use is not trivial despite having some tracking solutions in place; our statistics show that LaMachine was installed at least over a thousand times, on
machines in over 25 different countries, since its inception\footnote{https://applejack.science.ru.nl/lamastats/lamachinestats.html}. In early 2018 version 2 was released which was a significant redesign, powered
by Ansible\footnote{\url{https://www.ansible.com}}. We ourselves use LaMachine as a means by to make our full software
stack available to end-users, whether those be individual researchers, research groups, or hosting providers like CLARIN
centres. LaMachine has in fact been adopted in all of those areas. We also see a regular influx of questions on our issue tracker, indicating community interest.

@@

In LaMachine, software is grouped into various ``packages'', each package\footnote{For those familiar with
Ansible, a package in LaMachine is an Ansible role} groups one or multiple programs that have some kind of relation. The
installation manifest lists all packages that will be installed, at the user's discretion. After the initial
installation, the user can always add more packages using \texttt{lamachine-add} or by editing the installation manifest directly.

@@

In any of its flavours, LaMachine offers low-level shell access, i.e. a command-line interface accessed through a
terminal. In flavours that are separated from the host system by a network, this is accomplished over \texttt{ssh}.
After accessing the LaMachine environment through a terminal, as shown in Figure~\ref{fig:venv}, the user has the liberty to do whatever she wants and can invoke any
of the tools that offer a text interface, including text editors such as vim, emacs or nano, version control systems
such as git and interpreters such as Python or R (or compilers for that matter). This allow users to use
one of the many specialised programming libraries included in LaMachine to build their own tools. All this makes
LaMachine ideally suited as a development environment.

@@

LaMachine includes various CLAM-based webservices. CLAM is a long-time CLARIN project that allows developers to turn
their command-line NLP tool into a RESTful webservice with relatively little effort \cite{CLAMPAPER}, and which in
addition also provides a generic web-interface for human end-users. The portal website in LaMachine links to these. This
again addresses an audience that does not require a high degree of technical expertise, but is suitable for a wider
audience of researchers, including those in the Humanities.

@@

Within the scope of CLARIN, one might be inclined to draw comparisons with the CLARIN Language Resource Switchboard
\cite{zinn2016clarin}, which offers access to NLP tools through a web interface where users can upload their texts for
NLP processing, providing easy access and automated tool discovery. This, however, is notably different from what
LaMachine is. LaMachine is a software distribution; the CLARIN switchboard is not. Vice versa, LaMachine does not
currently provide any switchboard functionality, it merely offers a portal interface. The switchboard allows access to
tools hosted elsewhere, LaMachine installs and configures tools and only provides access to those it installed. Unlike
the switchboard, the audience for LaMachine is in the first place the data scientist and researcher who has
programming experience and wants to work with complex NLP software, on her own machine and unencumbered by high-level
interfaces if she so desires.

There are also other CLARIN initiatives that provide not just a switchboard, but aim at an eve more ambitious high-level virtual research
environment (VRE) with graphical interfaces to accommodate the researcher and allow execution and visualisation of
various NLP pipeline. The Dutch CLARIAH WP3 project\footnote{\url{https://github.com/meertensinstituut/clariah-wp3-vre}}
is such an initiative: LaMachine is currently developed in close collaboration with this project, in a role of technology supplier.
